Whole-body connectome of a segmented annelid larva

### 2024-08-16 Manuscript.qmd
Revision: change outgoing, incoming to post- presyn in figs and text, add refs

---
title: "Whole-body connectome of a segmented annelid larva"
subtitle: |
  \raggedright  Csaba Verasztó^1,2\†^, Sanja Jasek^1,3,\†^, Martin Gühmann^4^, Luis Alberto Bezares-Calderón^1^, Elizabeth A. Williams^5^, Réza Shahidi^6^, Gáspár Jékely^1,3,\*^  \newline<br>^1^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom \newline<br>^2^École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland, \newline<br>^3^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany \newline<br>^4^School of Biological Sciences, University of Bristol, Bristol, United Kingdom \newline<br>^5^BioSciences, University of Exeter, Stocker Road, Exeter EX4 4QD, Exeter, United Kingdom\newline<br> ^6^Electron Microscopy Core Facility (EMCF), University of Heidelberg, 69120 Heidelberg, Germany \newline<br> ^\*^Correspondence: gaspar.jekely@cos.uni-heidelberg.de \newline<br> ^\†^These authors contributed equally to this work
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
crossref:
  fig-title: Figure     # (default is "Figure")
max-width: 100%
format:
  html: default
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

```{r, out.width = "100%", fig.cap="**Figure 1. Overview of nervous-system anatomy in the three-day-old *Platynereis* larva.** (A) Stylised scanning electron microscopy image of a three-day-old segmented *Platynereis* larva. (B) Morphological rendering of all cells in the EM volume. Spheres represent the position of cell somas. (C) Neurite processes of all neurons in the larva coloured by neuron type (ventral view). Soma are not shown. (D) All effector cells in the larva, including ciliated cells (yellow), glands (grey) and muscles (red). Spheres indicate cell soma (not shown for muscle cells) (E) Summary of cell numbers of different categories in the larva.  (F-H) All presynaptic sites coloured by neuron type shown in (F) ventral, lateral (G), and frontal (H) views. (I) Morphological rendering of all sensory and interneurons on the left side of the larva.  (J) Morphological rendering of all sensory and motor neurons on the left side of the larva. Horizontal lines indicate the position of the cross-sections in (K). (K) Cross-section view of sensory, inter- and motor neuron projections in the ventral nerve cord neuropil. Cross-sections at three antero-posterior positions are shown. Numbers indicate the position of the cross-sections as marked in (J). (L) Position of presynaptic and postsynaptic sites on all sensory neurons of the connectome. (M) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in sensory neurons. (N) Position of presynaptic and postsynaptic sites on all interneurons of the connectome. (O) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in interneurons. (P) Position of presynaptic and postsynaptic sites on all motor neurons of the connectome. (Q) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in motor neurons. In all anatomical renderings, the outline of the yolk is shown in grey. In (G), the body outline is also shown. Aciculae and chaetae are also shown in grey as segmental markers. Abbreviations: SN, sensory neuron; IN, interneuron; MN, motor neuron, sg0-3, segments 0-3. <br> Figure 1—source data 1. Source data for panels M, O, Q. " }
#| echo: false

@@

Neurons have a median of `r median_presyn_sites` outgoing (presynpatic) and `r median_postsyn_sites` incoming (postsynaptic) synaptic sites with the largest neurons (MC, Ser-tr1, Loop) with over 300 outgoing synapses (Figure 1---figure supplement 2).

The relative ratio of incoming and outgoing sites distinguishes sensory, inter- and motoneurons, with sensory neurons having relatively more outgoing synapses (Figure 1---figure supplement 2E).

The number of outgoing and incoming synaptic sites and skeleton segments scales with the total cable length of skeletons with Pearson's correlation coefficients \>0.7 (Figure 1---figure supplement 2F-H).

Most neuronal trees are unipolar and not highly branched (<100 segments; Figure 1---figure supplement 2H).
Neurons in general do not have primary cilia.
Sensory neurons are bipolar with an axon and a sensory dendrite bearing zero to five sensory cilia at its proximal tip, showing different sensory specialisations [@williams2017; @bezares-calderon2018].
Interneurons are unipolar or pseudo-unipolar.
Ciliomotor neurons have two motor axons emanating from the soma [@veraszto2017], other motoneurons are unipolar.

@@

Outgoing and incoming synapses in a neuron can be intermingled or spatially segregated.
Interneurons can have mixed or spatially segregated input-output compartments. Most motoneurons have spatially segregated input-output compartments (Figure 1L-Q and Figure 3---figure supplement 7, 8).

@@

To explore sensory-motor pathways in the grouped graph, we searched for the shortest directed paths from all sensory neurons to all effector cells.
We identified four sensory neurons with direct motor output (sensory-motor neurons: eyespot-PRCR3, pygPBunp, pygCirrusPUSM, hPU2l-asymPDF) and 25 premotor sensory neurons.
The maximum number of hops from sensory neurons to an effector is five.
33 sensory neurons have no synaptic paths to effectors.
19 of these are neurosecretory cells with few or no synaptic partners [@williams2017] (Figure 4---figure supplement 5). For a graph-based visualisation of information flow, we arranged the nodes according to their relative ratio of incoming and outgoing synapses (Figure 4---figure supplement 6). This layout indicates that sensory-evoked activity propagates from sensory neurons through interneurons and motoneurons to effectors.

@@

To identify neurons with possible integrative function, we ranked cells by the number of presynaptic and postsynaptic cell-type-level partners. We also ranked cells by various node-centrality measures including node weighted degree, pagerank, betweenness and authority (Figure 7J, K; Figure 7---figure supplement 2).

@@

The overall architecture of the _Platynereis_ MB is multilayer, parallel and feed-forward, with only very weak recurrent connections or lateral connections between the feed-forward networks (Figure 9B; Figure 9---figure supplement 1). The circuit also shows a fan-out fan-in architecture, with the sensory neurons diverging to several IN targets (e.g. SNtorii to INtorii, INbigloop and INhorn) and post-MB interneurons receiving converging input (e.g. INW from INhook, INlasso, INMBtype5-6, and SNlasso). This multilayer perceptron architecture may support the formation of associative memories, as proposed for memory circuits in cephalopods [@shomrat2011alternative].

@@

The trunk-projecting brain neurons have either a decussating or ipsilaterally descending morphology.
The two types project to different domains of the VNC and have different target neurons (Figure 10---figure supplement 1E, F).
Decussating and descending neurons receive most of their incoming synapses in the central head neuropil and form most of their outgoing synapses in the VNC (Figure 10---figure supplement 1C, D, H-M).

@@

```{r, out.width = "100%", fig.cap="**Figure 15. Chaetal mechanosensors and their circuits.** (A) Morphological rendering of chaeMech neurons, ventral view. (B) Outgoing (red) and incoming (cyan) synapses of chaeMech neurons, ventral view. The soma of the chaeMech cells is shown in grey. (C) Lateral view of chaeMech  neurons. (D) TEM image of the sensory dendrites of chaeMech neurons (yellow highlight) surrounding the chaetae (red highlight). (E) Grouped synaptic connectivity matrix of chaeMech, CR and SNbronto neurons and their downstream targets. (F) Same connectivity information as in (E) represented as a network. In (A-C) the yolk outline, stomodeum and the mechanosensory girdle neuropil are shown in grey for reference. In (A, C) the aciculae and chaetae are also shown. <br> Figure 15—source data 1. Connectivity matrix for the network in panels E, F. "}
#| echo: false
knitr::include_graphics("Figures/Figure15.png")
```

@@

### A whole-body resource for _Platynereis_

Here we described a whole-body connectome for the segmented three-day-old larva of the marine annelid _Platynereis dumerilii_.
_Platynereis_ is the third species for which such a whole-body resource is available, after _C. elegans_ and _C. intestinalis_  [@white1986structure; @cook2019whole; @ryan2016cns]. 
The power of such whole-body approaches lies in their comprehensive nature encompassing not only the nervous system but also all effectors and other cells. Overall, we identified, annotated and spatially mapped 294 cell types and a total of `r num_soma` cells.

_Platynereis_ is distinguished from the nematode and tunicate larval connectomes by the complexity of its nervous and effector systems. The nervous system contains an order of magnitude more neurons. The effector system is multi-modal, and includes muscles, locomotor multiciliated cells, glands and pigment cells. The musculature is composed of 53 distinct muscle cell-types, is segmental, and extends to the parapodia that are supported by a chitin-based endoskeleton (aciculae) [@jasek2022].

@@

The mushroom bodies also show a parallel feed-forward organisation with minimal feedback. This could be due to the developmental snapshot we acquired for the three-day-old stage. Mushroom bodies grow in size and form a morphologically clearly recognisable region only in later-stage larvae and juveniles [@tomer2010profiling]. This growth is likely accompanied by circuit maturation. However, we think that the mushroom body cell types and circuits we reconstructed already show a functional circuit architecture. The MB cell types and their connections show left-right stereotypy and specificity of connectivity. It may be that annelid mushroom bodies can support associative learning by this multilayer perceptron-like organisation. In the cephalopods _Sepia officinalis_ and _Octopus vulgaris_, similar feedforward information flow characterises the vertical lobe, a learning centre  [@shomrat2011alternative]. The circuitry here has a simple fan-out fan-in architecture that in _Octopus vulgaris_ shows further interconnections between the parallel feedforward networks [@bidel2023connectomics]. In _Platynereis_, we only identified very weak connections between the parallel networks, at the level of the MB projection neurons.

@@

By comprehensive tracing and cell annotation, we also identified a unique class of pigment-motor neurons (MC3cover, cioMNcover, pygPBunp). The direct synaptic innervation of pigment cells suggests that pigmentation may be under neuronal control in _Platynereis_ larvae. We do not know how neuronal inputs may influence pigment cells. One possibility is that synaptic signals induce the movement or bleaching of the pigment vacuoles. In the cover cells, we occasionally observed the rapid disappearance of the pigment granules. The fact that at least one of the pigment-motor cells (pygPBunp) is also directly mechanosensory [@bezares-calderon2018] suggests that mechanical cues may alter pigmentation.

@@

The *Platynereis* whole-body connectome resource highlights the value of having access to comprehensive circuit and cell-type information. In the _Platynereis_ larva, almost every functional module spans several body regions and often the entire body.

@@

An exciting perspective in connectomics is to learn about circuit evolution by the analysis of comprehensive datasets that integrate anatomy and connectivity. 
Our reconstructions identified a potential case for circuit evolution by duplication and divergence of synaptically connected cell types [@tosches2017developmental].

@@

The duplication and divergence of cell-type sets also characterised the evolution of the vertebrate cerebellum [@kebschull2020cerebellar].
Further support for this model in *Platynereis* could come from comprehensive gene expression analyses for the distinct mechanosensory and INsplit neurons.

@@

We identified synapses based on a vesicle cloud close to the plasma membrane.
Most synapses were visible in consecutive layers (for example images see [@randel2014neuronal] and browse the data).
We also checked for the proximity of mitochondria in the same arbour in case of ambiguous synapses --- a requirement supported by quantitative connectomic data in _Drosophila_ [@Schneider-Mizell2016].
The systematic review of all neurons belonging to a cell type was done by one or multiple reviewers until close to 100% was reached for every cell.
Cells were further checked in the 3D widget to split implausible skeletons.
Synapses were reviewed multiple times, from both the pre- and postsynaptic arbour.

@@

All cells have a unique name.
We named neurons based on their type (e.g. sensory or motor; SN, MN) cell body position (left or right, head or trunk segment), axonal morphology, neuropeptide expression, and other specialisations (e.g. sensory morphology).
Cells of the same type have similar names, distinguished by body position indicators and numbers. These indicators follow the general cell-type name separated by the first _ symbol in the name string.
We endeavoured to give names that were easy to remember.
The name of many sensory neurons start with SN followed by a specific term (e.g. blunt, bronto, stiff).
Interneurons often start with IN and motoneurons with MN.
There are exceptions, including neurons with known function (e.g. PRC for photoreceptor cells) and neurons with prominent morphology (e.g. Loop).
Segmental position (sg0-3) and body side (l or r) is indicated in the name of most neurons.
Non-neuronal cells were named based on anatomical terms (e.g. prototroch) or by abbreviations (e.g. EC for epithelial cell).

All cells have multiple annotations, which can be used to query the database in CATMAID or through the CATMAID API (e.g. in R b the catmaid package).
Neurons belonging to a cell-type category were annotated with the generalist annotation 'celltype' and a cell-type-specific annotation (e.g. celltype23 for the INpreMN neurons).
Non-neuronal cells belonging to a cell-type category were annotated with the generalist annotation 'celltype_non_neuronal' and a cell-type-specific annotation (e.g. celltype_non_neuronal23 for the acicula cells).
Neurons were also annotated with descriptors of their projection morphologies (e.g. commissural, ipsilateral, pseudounipolar etc.), neuron class (Sensory neuron, sensory-motor neuron, sensory-neurosecretory neuron, interneuron, inter-motorneuron and motorneuron [note the 'r', which we kept for backward compatibility]).
In CATMAID, we recommend the use of regular expressions for searching annotations e.g. \^motorneuron\$ to retrieve exact matches.
Differentiating neurons with immature sensory dendrites or axonal projections with axonal growth cones and with no or few synapses were annotated 'immature neuron' (402 cells).
Ascending trunk neurons and descending head neurons traversing the circumesophageal connectives were annotated 'head-trunk'.
Neurons with a soma in the head and a descending decussating axon were annotated with 'decussating'.
Cells were also annotated according to the location of their soma in a certain body region (head --- as 'episphere', trunk --- as 'torso', 'pygidium'), body side ('left_side', 'right_side'), segment ('segment_0' etc.), and germ layer (ecto-, meso-, endoderm).

@@

When defining the connectome, we aimed at including differentiated cells and skeletons connected with at least three synapses to the main graph.
We used the script 'connectome_from_CATMAID.R' to derive the final full connectome graph.
First, we fetched all synaptic connectors and their pre- and postsynaptoc partners. Each single synapse was assigned a weight of one and edges connecting the same nodes in the same direction were summed. We then removed all vertices from the graph with <3 synapses. We checked for connected components and selected the largest subgraph.

@@

```{r, out.width = "100%", fig.cap="**Figure 3—figure supplement 7. Radial density of incoming and outgoing synapses in head neurons. ** (A-T) Morphological renderings of selected head cell types with input and output synapses (left panels) and radial density plots of incoming and outgoing synapses (right panels). Synapse density was calculated from the root node (soma) and was averaged for different neurons of the same cell type. <br> Figure3—figure supplement 7—source data 1. Radial density data of incoming and outgoing synapses for all 202 cell types. "}
#| echo: false
knitr::include_graphics("Figures/Figure3_fig_suppl7.png")
```

```{r, out.width = "100%", fig.cap="**Figure 3—figure supplement 8. Radial density of incoming and outgoing synapses in trunk neurons. ** (A-T) Morphological renderings of selected trunk cell types with input and output synapses (left panels) and radial density plots of incoming and outgoing synapses (right panels). Synapse density was calculated from the root node (soma) and was averaged for different neurons of the same cell type. Source data are the same as for Figure3—figure supplement 7. "}
#| echo: false
knitr::include_graphics("Figures/Figure3_fig_suppl8.png")
```

@@

```{r, out.width = "100%", fig.cap="**Figure 10—figure supplement 1. Head descending and decussating neurons.** (A) Morphological rendering of the neurites of head desending (blue) and decussationg (yellow) neurons. (B) Names of head descending and decussationg cell types. (C) Incoming synapses on head descending and decussating neurons. (D) Outgoing synapses of head descending and decussating neurons. (E) Postsynaptic targets of head descending and decussating neurons. (F) Distribution of the neurites of head descending and decussating neurons in the VNC. Cross section at the position of the thin line in (E). (G) Neurites of the postsynaptic targets of head descending and decussating neurons in the VNC. Cross section at the position of the thin line in (E). (H-I) Incoming (red) and outgoing (blue) synapses of head decussating neurons in anterior (H) and ventral (I) view. (J) Mean radial density of incoming and outgoing synapses in head decussating neurons. (K-L) Incoming (red) and outgoing (blue) synapses of head descending neurons in anterior (K) and ventral (L) view. (M) Mean radial density of incoming and outgoing synapses in head descending neurons. <br> Figure 10—figure supplement 1—source data 1. Source data for panels (J, M). "}
#| echo: false
knitr::include_graphics("Figures/Figure10_fig_suppl1.png")
```

@@

# References
\

>>>>

---
title: "Whole-body connectome of a segmented annelid larva"
subtitle: |
  \raggedright  Csaba Verasztó^1,2\†^, Sanja Jasek^1,3,\†^, Martin Gühmann^4^, Luis Alberto Bezares-Calderón^1^, Elizabeth A. Williams^5^, Réza Shahidi^6^, Gáspár Jékely^1,3,\*^  \newline<br>^1^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom \newline<br>^2^École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland, \newline<br>^3^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany \newline<br>^4^School of Biological Sciences, University of Bristol, Bristol, United Kingdom \newline<br>^5^BioSciences, University of Exeter, Stocker Road, Exeter EX4 4QD, Exeter, United Kingdom\newline<br> ^6^Electron Microscopy Core Facility (EMCF), University of Heidelberg, 69120 Heidelberg, Germany \newline<br> ^\*^Correspondence: gaspar.jekely@cos.uni-heidelberg.de \newline<br> ^\†^These authors contributed equally to this work
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
crossref:
  fig-title: Figure     # (default is "Figure")
max-width: 100%
format:
  html: default
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
editor: 
  markdown: 
    wrap: sentence
---

@@

```{r, out.width = "100%", fig.cap="**Figure 1. Overview of nervous-system anatomy in the three-day-old *Platynereis* larva.** (A) Stylised scanning electron microscopy image of a three-day-old segmented *Platynereis* larva. (B) Morphological rendering of all cells in the EM volume. Spheres represent the position of cell somas. (C) Neurite processes of all neurons in the larva coloured by neuron type (ventral view). Soma are not shown. (D) All effector cells in the larva, including ciliated cells (yellow), glands (grey) and muscles (red). Spheres indicate cell soma (not shown for muscle cells) (E) Summary of cell numbers of different categories in the larva.  (F-H) All presynaptic sites coloured by neuron type shown in (F) ventral, lateral (G), and frontal (H) views. (I) Morphological rendering of all sensory and interneurons on the left side of the larva.  (J) Morphological rendering of all sensory and motor neurons on the left side of the larva. Horizontal lines indicate the position of the cross-sections in (K). (K) Cross-section view of sensory, inter- and motor neuron projections in the ventral nerve cord neuropil. Cross-sections at three antero-posterior positions are shown. Numbers indicate the position of the cross-sections as marked in (J). (L) Position of presynaptic and postsynaptic sites on all sensory neurons of the connectome. (M) Mean radial synapse density (radius: 1000 nm, centre: soma) of presynaptic and postsynaptic synapses in sensory neurons. (N) Position of presynaptic and postsynaptic sites on all interneurons of the connectome. (O) Mean radial synapse density (radius: 1000 nm, centre: soma) of presynaptic and postsynaptic synapses in interneurons. (P) Position of presynaptic and postsynaptic sites on all motor neurons of the connectome. (Q) Mean radial synapse density (radius: 1000 nm, centre: soma) of presynaptic and postsynaptic synapses in motor neurons. In all anatomical renderings, the outline of the yolk is shown in grey. In (G), the body outline is also shown. Aciculae and chaetae are also shown in grey as segmental markers. Abbreviations: SN, sensory neuron; IN, interneuron; MN, motor neuron, sg0-3, segments 0-3. <br> Figure 1—source data 1. Source data for panels M, O, Q. " }
#| echo: false

@@

Neurons have a median of `r median_presyn_sites` presynpatic and `r median_postsyn_sites` postsynaptic sites with the largest neurons (MC, Ser-tr1, Loop) with over 300 presynapses (Figure 1---figure supplement 2).

The relative ratio of post- and presynaptic sites distinguishes sensory, inter- and motoneurons, with sensory neurons having relatively more presynapses (Figure 1---figure supplement 2E).

The number of pre- and postsynaptic sites and skeleton segments scales with the total cable length of skeletons with Pearson's correlation coefficients \>0.7 (Figure 1---figure supplement 2F-H).

Most neuronal trees are unipolar and not highly branched (\<100 segments; Figure 1---figure supplement 2H).
Neurons in general do not have primary cilia.
Sensory neurons are bipolar with an axon and a sensory dendrite bearing zero to five sensory cilia at its proximal tip, showing different sensory specialisations [@williams2017; @bezares-calderon2018].
Interneurons are unipolar or pseudo-unipolar.
Ciliomotor neurons have two motor axons emanating from the soma [@veraszto2017], other motoneurons are unipolar.

@@

Presynaptic and postsynaptic sites in a neuron can be intermingled or spatially segregated.
Interneurons can have mixed or spatially segregated input-output compartments.
Most motoneurons have spatially segregated input-output compartments (Figure 1L-Q and Figure 3---figure supplement 7, 8).

@@

To explore sensory-motor pathways in the grouped graph, we searched for the shortest directed paths from all sensory neurons to all effector cells.
We identified four sensory neurons with direct motor output (sensory-motor neurons: eyespot-PRCR3, pygPBunp, pygCirrusPUSM, hPU2l-asymPDF) and 25 premotor sensory neurons.
The maximum number of hops from sensory neurons to an effector is five.
33 sensory neurons have no synaptic paths to effectors.
19 of these are neurosecretory cells with few or no synaptic partners [@williams2017] (Figure 4---figure supplement 5).
For a graph-based visualisation of information flow, we arranged the nodes according to their relative ratio of post- and presynapses (Figure 4---figure supplement 6).
This layout indicates that sensory-evoked activity propagates from sensory neurons through interneurons and motoneurons to effectors.

@@

To identify neurons with possible integrative function, we ranked cells by the number of presynaptic and postsynaptic cell-type-level partners.
We also ranked cells by various node-centrality measures including node weighted degree, PageRank, betweenness and authority (Figure 7J, K; Figure 7---figure supplement 2). PageRank is an algorithm used to rank webpages and scores the number and quality of the incoming links of a node [@page1999pagerank], betweenness centrality measures the number of shortest paths that pass through a node in a graph,  and authority [@kleinberg1999authoritative].

@@

The overall architecture of the *Platynereis* MB is multilayer, parallel and feed-forward, with only very weak recurrent connections or lateral connections between the feed-forward networks (Figure 9B; Figure 9---figure supplement 1).
The circuit also shows a fan-out fan-in architecture, with the sensory neurons diverging to several IN targets (e.g. SNtorii to INtorii, INbigloop and INhorn) and post-MB interneurons receiving converging input (e.g. INW from INhook, INlasso, INMBtype5-6, and SNlasso).
This multilayer perceptron architecture may support the formation of associative memories, as proposed for memory circuits in cephalopods [@shomrat2011alternative].

@@

The trunk-projecting brain neurons have either a decussating or ipsilaterally descending morphology.
The two types project to different domains of the VNC and have different target neurons (Figure 10---figure supplement 1E, F).
Decussating and descending neurons receive most of their synaptic inputs (postsynaptic sites) in the central head neuropil and form most of their presynapses in the VNC (Figure 10---figure supplement 1C, D, H-M).

@@

```{r, out.width = "100%", fig.cap="**Figure 15. Chaetal mechanosensors and their circuits.** (A) Morphological rendering of chaeMech neurons, ventral view. (B) Presynaptic (red) and postsynaptic (cyan) sites of chaeMech neurons, ventral view. The soma of the chaeMech cells is shown in grey. (C) Lateral view of chaeMech  neurons. (D) TEM image of the sensory dendrites of chaeMech neurons (yellow highlight) surrounding the chaetae (red highlight). (E) Grouped synaptic connectivity matrix of chaeMech, CR and SNbronto neurons and their downstream targets. (F) Same connectivity information as in (E) represented as a network. In (A-C) the yolk outline, stomodeum and the mechanosensory girdle neuropil are shown in grey for reference. In (A, C) the aciculae and chaetae are also shown. <br> Figure 15—source data 1. Connectivity matrix for the network in panels E, F. "}
#| echo: false
knitr::include_graphics("Figures/Figure15.png")
```

@@

### A whole-body resource for *Platynereis*

Here we described a whole-body connectome for the segmented three-day-old larva of the marine annelid *Platynereis dumerilii*.
*Platynereis* is the third species for which such a whole-body resource is available, after *C. elegans* and *C. intestinalis* [@white1986structure; @cook2019whole; @ryan2016cns].
The power of such whole-body approaches lies in their comprehensive nature encompassing not only the nervous system but also all effectors and other cells.
Overall, we identified, annotated and spatially mapped 294 cell types and a total of `r num_soma` cells.

*Platynereis* is distinguished from the nematode and tunicate larval connectomes by the complexity of its nervous and effector systems.
The nervous system contains an order of magnitude more neurons.
The effector system is multi-modal, and includes muscles, locomotor multiciliated cells, glands and pigment cells.
The musculature is composed of 53 distinct muscle cell-types, is segmental, and extends to the parapodia that are supported by a chitin-based endoskeleton (aciculae) [@jasek2022].

@@

The mushroom bodies also show a parallel feed-forward organisation with minimal feedback.
This could be due to the developmental snapshot we acquired for the three-day-old stage.
Mushroom bodies grow in size and form a morphologically clearly recognisable region only in later-stage larvae and juveniles [@tomer2010profiling].
This growth is likely accompanied by circuit maturation.
However, we think that the mushroom body cell types and circuits we reconstructed already show a functional circuit architecture.
The MB cell types and their connections show left-right stereotypy and specificity of connectivity.
It may be that annelid mushroom bodies can support associative learning by this multilayer perceptron-like organisation.
In the cephalopods *Sepia officinalis* and *Octopus vulgaris*, similar feedforward information flow characterises the vertical lobe, a learning centre [@shomrat2011alternative].
The circuitry here has a simple fan-out fan-in architecture that in *Octopus vulgaris* shows further interconnections between the parallel feedforward networks [@bidel2023connectomics].
In *Platynereis*, we only identified very weak connections between the parallel networks, at the level of the MB projection neurons.

@@

By comprehensive tracing and cell annotation, we also identified a unique class of pigment-motor neurons (MC3cover, cioMNcover, pygPBunp).
The direct synaptic innervation of pigment cells suggests that pigmentation may be under neuronal control in *Platynereis* larvae.
We do not know how neuronal inputs may influence pigment cells.
One possibility is that synaptic signals induce the movement or bleaching of the pigment vacuoles.
In the cover cells, we occasionally observed the rapid disappearance of the pigment granules.
The fact that at least one of the pigment-motor cells (pygPBunp) is also directly mechanosensory [@bezares-calderon2018] suggests that mechanical cues may alter pigmentation.

@@

The *Platynereis* whole-body connectome resource highlights the value of having access to comprehensive circuit and cell-type information.
In the *Platynereis* larva, almost every functional module spans several body regions and often the entire body.

@@

An exciting perspective in connectomics is to learn about circuit evolution by the analysis of comprehensive datasets that integrate anatomy and connectivity.
Our reconstructions identified a potential case for circuit evolution by duplication and divergence [@tosches2017developmental; @roberts2022evolution], a concept that originated from genetics [@ohno1970evolution].

@@

The duplication and divergence of cell-type sets also characterised the evolution of the vertebrate cerebellum [@kebschull2020cerebellar].
Duplication and divergence of circuits was also proposed as a possible mechanism for the evolution of brain pathways for vocal learning in song-learning birds, spoken language in humans [@chakraborty2015brain] and other circuits [@roberts2022evolution]. Further support for this model in *Platynereis* could come from comprehensive gene expression analyses for the distinct mechanosensory and INsplit neurons.

@@

We identified synapses based on a vesicle cloud close to the plasma membrane.
Most synapses were visible in consecutive layers (for example images see [@randel2014neuronal] and browse the data).
We also checked for the proximity of mitochondria in the same arbour in case of ambiguous synapses --- a requirement supported by quantitative connectomic data in *Drosophila* [@Schneider-Mizell2016].
The systematic review of all neurons belonging to a cell type was done by one or multiple reviewers until close to 100% was reached for every cell.
Cells were further checked in the 3D widget to split implausible skeletons.
Synapses were reviewed multiple times, from both the pre- and postsynaptic arbour.

@@

All cells have a unique name.
We named neurons based on their type (e.g. sensory or motor; SN, MN) cell body position (left or right, head or trunk segment), axonal morphology, neuropeptide expression, and other specialisations (e.g. sensory morphology).
Cells of the same type have similar names, distinguished by body position indicators and numbers.
These indicators follow the general cell-type name separated by the first \_ symbol in the name string.
We endeavoured to give names that were easy to remember.
The name of many sensory neurons start with SN followed by a specific term (e.g. blunt, bronto, stiff).
Interneurons often start with IN and motoneurons with MN.
There are exceptions, including neurons with known function (e.g. PRC for photoreceptor cells) and neurons with prominent morphology (e.g. Loop).
Segmental position (sg0-3) and body side (l or r) is indicated in the name of most neurons.
Non-neuronal cells were named based on anatomical terms (e.g. prototroch) or by abbreviations (e.g. EC for epithelial cell).

All cells have multiple annotations, which can be used to query the database in CATMAID or through the CATMAID API (e.g. in R b the catmaid package).
Neurons belonging to a cell-type category were annotated with the generalist annotation 'celltype' and a cell-type-specific annotation (e.g. celltype23 for the INpreMN neurons).
Non-neuronal cells belonging to a cell-type category were annotated with the generalist annotation 'celltype_non_neuronal' and a cell-type-specific annotation (e.g. celltype_non_neuronal23 for the acicula cells).
Neurons were also annotated with descriptors of their projection morphologies (e.g. commissural, ipsilateral, pseudounipolar etc.), neuron class (Sensory neuron, sensory-motor neuron, sensory-neurosecretory neuron, interneuron, inter-motorneuron and motorneuron \[note the 'r', which we kept for backward compatibility\]).
In CATMAID, we recommend the use of regular expressions for searching annotations e.g. \^motorneuron\$ to retrieve exact matches.
Differentiating neurons with immature sensory dendrites or axonal projections with axonal growth cones and with no or few synapses were annotated 'immature neuron' (402 cells).
Ascending trunk neurons and descending head neurons traversing the circumesophageal connectives were annotated 'head-trunk'.
Neurons with a soma in the head and a descending decussating axon were annotated with 'decussating'.
Cells were also annotated according to the location of their soma in a certain body region (head --- as 'episphere', trunk --- as 'torso', 'pygidium'), body side ('left_side', 'right_side'), segment ('segment_0' etc.), and germ layer (ecto-, meso-, endoderm).

@@

When defining the connectome, we aimed at including differentiated cells and skeletons connected with at least three synapses to the main graph.
We used the script 'connectome_from_CATMAID.R' to derive the final full connectome graph.
First, we fetched all synaptic connectors and their pre- and postsynaptoc partners.
Each single synapse was assigned a weight of one and edges connecting the same nodes in the same direction were summed.
We then removed all vertices from the graph with \<3 synapses.
We checked for connected components and selected the largest subgraph.

@@

```{r, out.width = "100%", fig.cap="**Figure 3—figure supplement 7. Radial density of presynaptic and postsynaptic sites in head neurons. ** (A-T) Morphological renderings of selected head cell types with input and output synapses (left panels) and radial density plots of presynaptic (blue) and postsynaptic (orange) sites (right panels). Synapse density was calculated from the root node (soma) and was averaged for different neurons of the same cell type. <br> Figure3—figure supplement 7—source data 1. Radial density data of incoming (postsynaptic) and outgoing (presynaptic) synapses for all 202 cell types. "}
#| echo: false
knitr::include_graphics("Figures/Figure3_fig_suppl7.png")
```

```{r, out.width = "100%", fig.cap="**Figure 3—figure supplement 8. Radial density of presynaptic and postsynaptic sites  in trunk neurons. ** (A-T) Morphological renderings of selected trunk cell types with input and output synapses (left panels) and radial density plots of presynaptic (blue) and postsynaptic (orange) sites (right panels). Synapse density was calculated from the root node (soma) and was averaged for different neurons of the same cell type. Source data are the same as for Figure3—figure supplement 7. "}
#| echo: false
knitr::include_graphics("Figures/Figure3_fig_suppl8.png")
```

@@

```{r, out.width = "100%", fig.cap="**Figure 10—figure supplement 1. Head descending and decussating neurons.** (A) Morphological rendering of the neurites of head desending (blue) and decussationg (yellow) neurons. (B) Names of head descending and decussationg cell types. (C) Postsynaptic sites of head descending and decussating neurons. (D) Presynaptic sites of head descending and decussating neurons. (E) Postsynaptic targets of head descending and decussating neurons. (F) Distribution of the neurites of head descending and decussating neurons in the VNC. Cross section at the position of the thin line in (E). (G) Neurites of the postsynaptic targets of head descending and decussating neurons in the VNC. Cross section at the position of the thin line in (E). (H-I) Postsynaptic (red) and presynaptic (blue) sites of head decussating neurons in anterior (H) and ventral (I) view. (J) Mean radial density of postsynapses and presynapses in head decussating neurons. (K-L) Postsynaptic (red) and presynaptic (blue) sites of head descending neurons in anterior (K) and ventral (L) view. (M) Mean radial density of postsynapses and presynapses in head descending neurons. <br> Figure 10—figure supplement 1—source data 1. Source data for panels (J, M). "}
#| echo: false
knitr::include_graphics("Figures/Figure10_fig_suppl1.png")
```

@@

# References
